Reinforcement Learning for Options on Target Volatility Funds

### 2021-05-28 tex/targetvol.tex
Paper
 first completed version

\date{
	First Version: March 2021. This Version: \today
}
\begin{document} % goes here

@@

 \subsection{Episode Description: Agent-Environment Interaction}
 We consider an episode $\tau$ of length $m+1$ that takes place on a discrete time-grid of fixing times expressed in year fractions $\mathcal{T}:=\{T_0,...,T_k,...,T_m\}$ with $T_0:=0$ and $T_m=T$ maturity of the option. At a given episodic time $T_k$ the RL agent interacts with the environment: it receives a representation of the environment called state $s_k$ and on the basis of that it selects an action $a_k$ sampling from the current policy $\pi^\theta_k$.
 In our case the agent can choose the composition of the risky asset portfolio, so that the policy is the asset allocation weights $\alpha$. From the HJB equation we have derived in section \Cref{subsec:HJB}, we select as state of th environment the following block
 \begin{equation}
 	s_{k} \coloneqq \left[T_k,I_{T_k}, S_{T_k}\right] \quad \forall T_k \in \mathcal{T}.
 \label{eq:state}\end{equation}
 This state contains all the informations needed by the agent to take an optimal action, leading to the maximum option price on the TVS. 
 Once the agent has selected the action $a_k$, it receives the next time $T_{k+1}$ a reward $r_{k+1}$ generated by the environment. We have defined the reward function in two different ways. A first definition is
 \begin{equation}
 	r_{k+1}=\left\{\begin{array}{ll}
 		(I_T - K)^+ & \text { if } T_{k+1} = T \\
 		0 & \text { otherwise }
 	\end{array}\right.
 \end{equation}
Therefore, during the episode, the agent receives a nil reward except at maturity when the reward coincides with the option intrinsic value. This choice may seem too daring because the agent receives a real feedback of its actions only at the end of the whole episode, leading to a slow learning. However, once the agent has learnt the optimal policy $\pi^* = \pi^{\theta^*}$, the average cumulative reward per episodes will coincide to the optimal TVO price.
The second definition of reward function for this problem we have developed is
 \begin{equation}
 	r_{k+1} \coloneqq  \gamma^{k}[V_{\text{BS}}(T_{k+1})-V_\text{\text{BS}}(T_k)],
 \label{eq:reward}\end{equation}
where $\gamma \in[0,1]$ is an hyper-parameter of the PPO, while $V_{\text{BS}}(T_k)$ is the BS price as proxy of the residual option price defined by
\begin{equation}
	V_{\text{BS}}(T_k) \coloneqq BS(F^{\text{TVS}}(T_k,T;\alpha^*_{\text{BS}},K),K,T-T_k,\bar{\sigma},D(T_k,T;\zeta)) \quad \text{and} \quad V(T_0)=0
\end{equation} 
with $\alpha^*_{BS}$ the BS optimal strategy \eqref{eq:BS_optimal_strategy} calculated in the state $s_k$. This second definition of the reward function does not suffer of nil rewards at intermediate times of the episode.  The hyper-parameter $\gamma$ plays the role of a discount rate in the sense that, as $\gamma$ approaches to zero, the RL agent will tend to maximize immediate rewards while neglecting possible larger rewards in the future. Choosing\footnote{We refer to \textcite{gae} for a more detailed description of $\lambda$.} $\gamma,\lambda<1$ and introduces a bias in the final result.
If we take the cumulative reward per episode and set $\gamma =\lambda =1$ we obtain
\begin{equation}
	R(\tau) = \sum_{k=0}^{m-1} r_{k+1} \underset{\gamma=\lambda =1}{=} \sum_{k=0}^{m-1}[V_{\text{BS}}(T_{k+1})-V_\text{\text{BS}}(T_k)] = V_{BS}(T) = (I_T-K)^+,
\end{equation} 
which is equal to the intrinsic value of the option. This result does not depend on the definition of $V_{\text{BS}}$  $\forall T_k<T$, but we think that the closer $V_{\text{BS}}$ is to the value function, the easier the agent is in learning.

@@

\subsection{Black and Scholes: Hyper-Parameters Fine Tuning}
We use the BS environment as toy model to understand which parameters of the RL algorithm play key roles in the training and testing phase. We also study which action parameterization is most effective in the case of a long only allocation strategy $\alpha$. We consider a target volatility call option with the following contract details between bank and fund manager
\begin{equation}\nonumber
	I_0 = K = 1 \, \text{EUR}, \quad T=1 \, \text{yr}, \quad \bar{\sigma} = 5\%.
\end{equation}
We test the action parameterizations discussed in \Cref{subsec:policy_action} increasing the number $n$ of equities in the basket; we have evidence that the $(n-1)$-sphere parameterization  \eqref{eq:euler_angles} suffers of labelling problems when $n>2$. In fact if we consider a spherical case ($n=3$), \cref{eq:euler_angles} introduces an asymmetry in the action space (\Cref{fig:n_sphere_parameterization} on the left) that translates into the choice by the agent of sub-optimal policies during the training phase. Because of that we privilege the action parameterization through normalization \eqref{eq:normalized_action}. 
 In \Cref{fig:n_sphere_parameterization} on the right we provide the learning curve of a $5\times8$ ANN\footnote{We use the notation $L\times N$ to describe a FFNN with $L$ hidden layers each made up of $N$ neurons.} with $\tanh$ as activation function, $\gamma=1$, $c_1=0.7$ and $c_2=0$. This learning curve, just like all the ones we will show in this work, is the best result in term of reward on the last $10^7$ episodes among four simulations with different initial seed for the weights $\theta$. This is done since the objective function is not convex.

@@

\begin{figure}[ht!]
	\centering
	\includegraphics[height=6cm,width=7.2cm]{Activation_functions_Learning_curve_1x40_network_confront_activation_strategy_EUR_8x5_900dpi.pdf}
	\includegraphics[height=6cm,width=7.2cm]{Value_function_Learning_curve_1x40_beta_confront_varying_strategy_EUR_8x5_400dpi_compressed.pdf}
	\includegraphics[height=6cm,width=7.2cm]{Hidden_Layers_Learning_curve_3asset_varystrategy_window1000000_EUR_5e8timestep_comparison_900dpi_compressed.pdf}
	\caption{Fine tuning simulations: on the top-left learning curves with different activation functions, on the top-right grid search on the policy-value trade-off and on bottom-centre learning curves on ANNs architectures. On the horizontal axis the number of training episodes. The solid lines are the moving average of the realized rewards on the last $5\times 10^5$ episodes. The shadows represent the $98\%$ confidence intervals. The dash-dotted horizontal red lines are the conservative TVO prices computed according to the optimal risky portfolio composition through \cref{eq:bs_closed_solution}.}
	\label{fig:fine_tuning}
\end{figure}  
\clearpage
\subsection{Local Volatility Dynamics}
In this section we study the TVS control problem assuming a local volatility model for the dynamics of the risky assets. Thus in this case we have a diffusive term in \Cref{eq:Equity_process} that is a deterministic function both of time and state, i.e. the spot price, $\nu_t=\nu(t,S_t)$. This additional dependency of volatility makes the problem of finding the optimal strategy non-trivial; in fact if we consider the whole equity smiles then the second order term in the HBJ equation \eqref{eq:HJB_general} is non null and thus a closed formula for $\alpha^*$ is not available anymore. Because of that one must resort on Reinforcement Learning to recover the problem solution. Unlike BS model where the strategy depends only on time, in LV dynamics there are no unnecessary information provided by the state block in \Cref{eq:state} to take the optimal action.

@@

\section{Conclusion and Further Developments}
In this paper we described a non-trivial control problem related to derivative contracts on target volatility strategies. We considered a bank selling a call option to a fund manager as protection  on the capital
invested on the TVS. We showed how the presence of different funding costs coming from hedging the risky assets underlying the TVS, makes the bank solving a stochastic optimal control problem to adjust the price of the protection, since the bank strategy is not self-financing. We derived a closed form solution of the control problem in a Black and Scholes framework in two different ways: first by applying the Gy\"ongy  Lemma and then by writing the Hamilton-Jacobi-Bellman equation. We studied the problem both in the BS world and in the local volatility one where the solution is not available and thus numerical investigation is needed. We tackled the problem by means of the novel Reinforcement Learning techniques, such as the proximal policy optimization. We used the BS model, where the solution is \textit{a priori} known, as benchmark to perform a series of fine-tuning of the RL algorithm hyper-parameters, such as the artificial neural network architecture. Then we investigated the problem under LV dynamics. From our simulations we had evidence that under this model the RL agent learnt as optimal strategy the application path-wise of the BS solution. This topic is still under study since our results for the LV model do not provide a formal demonstration of that the problem solution is the path-wise BS one. For this reason the natural developments of this work is to compare our RL results with other standard techniques like the least-square Monte Carlo. If this comparison is positive, then it will provide robustness of our LV results.
\begin{comment}

>>>>

%\date{
%	First Version: March 2021. This Version: \today
%}
\begin{document} % goes here

@@

 \subsection{Episode Description: Agent-Environment Interaction}
 We consider an episode $\tau$ of length $m+1$ that takes place on a discrete time-grid of fixing times expressed in year fractions $\mathcal{T}:=\{T_0,...,T_k,...,T_m\}$ with $T_0:=0$ and $T_m=T$ maturity of the option. At a given episodic time $T_k$ the RL agent interacts with the environment: it receives a representation of the environment called state $s_k$ and on the basis of that it selects an action $a_k$ sampling from the current policy $\pi^\theta_k$.
 In our case the agent can choose the composition of the risky asset portfolio, so that the policy is the asset allocation weights $\alpha$. From the HJB equation we have derived in section \Cref{subsec:HJB}, we select as state of th environment the following block
 \begin{equation}
 	s_{k} \coloneqq \left[T_k,I_{T_k}, S_{T_k}\right] \quad \forall T_k \in \mathcal{T}.
 \label{eq:state}\end{equation}
 This state contains all the informations needed by the agent to take an optimal action, leading to the maximum option price on the TVS. 
 Once the agent has selected the action $a_k$, it receives the next time $T_{k+1}$ a reward $r_{k+1}$ generated by the environment. We have defined the reward function in two different ways. A first definition is
 \begin{equation}
 	r_{k+1}=\left\{\begin{array}{ll}
 		(I_T - K)^+ & \text { if } T_{k+1} = T \\
 		0 & \text { otherwise }
 	\end{array}\right.
 \label{eq:reward2}\end{equation}
Therefore, during the episode, the agent receives a nil reward except at maturity when the reward coincides with the option intrinsic value. This choice may seem too daring because the agent receives a real feedback of its actions only at the end of the whole episode, leading to a slow learning. However, once the agent has learnt the optimal policy $\pi^* = \pi^{\theta^*}$, the average cumulative reward per episodes will coincide to the optimal TVO price.
The second definition of reward function for this problem we have developed is
 \begin{equation}
 	r_{k+1} \coloneqq  \gamma^{k}[V_{\text{BS}}(T_{k+1})-V_\text{\text{BS}}(T_k)],
 \label{eq:reward}\end{equation}
where $\gamma \in[0,1]$ is an hyper-parameter of the PPO, while $V_{\text{BS}}(T_k)$ is the BS price as proxy of the residual option price defined by
\begin{equation}
	V_{\text{BS}}(T_k) \coloneqq BS(F^{\text{TVS}}(T_k,T;\alpha^*_{\text{BS}},K),K,T-T_k,\bar{\sigma},D(T_k,T;\zeta)) \quad \text{and} \quad V(T_0)=0
\end{equation} 
with $\alpha^*_{BS}$ the BS optimal strategy \eqref{eq:BS_optimal_strategy} calculated in the state $s_k$. This second definition of the reward function does not suffer of nil rewards at intermediate times of the episode.  The hyper-parameter $\gamma$ plays the role of a discount rate in the sense that, as $\gamma$ approaches to zero, the RL agent will tend to maximize immediate rewards while neglecting possible larger rewards in the future. Choosing\footnote{We refer to \textcite{gae} for a more detailed description of $\lambda$.} $\gamma,\lambda<1$ and introduces a bias in the final result.
If we take the cumulative reward per episode and set $\gamma =\lambda =1$ we obtain
\begin{equation}
	R(\tau) = \sum_{k=0}^{m-1} r_{k+1} \underset{\gamma=\lambda =1}{=} \sum_{k=0}^{m-1}[V_{\text{BS}}(T_{k+1})-V_\text{\text{BS}}(T_k)] = V_{BS}(T) = (I_T-K)^+,
\end{equation} 
which is equal to the intrinsic value of the option. This result does not depend on the definition of $V_{\text{BS}}$  $\forall T_k<T$, but we think that the closer $V_{\text{BS}}$ is to the value function, the easier the agent is in learning.

@@

\subsection{Black and Scholes: Hyper-Parameters Fine Tuning}
We use the BS environment as toy model to understand which parameters of the RL algorithm play key roles in the training and testing phase. We also study which action parameterization is most effective in the case of a long only allocation strategy $\alpha$. We consider a target volatility call option with the following contract details between bank and fund manager
\begin{equation}\nonumber
	I_0 = K = 1 \, \text{EUR}, \quad T=1 \, \text{yr}, \quad \bar{\sigma} = 5\%.
\end{equation}
The reward function adopted in this environment is that expressed by \Cref{eq:reward2}. We test the action parameterizations discussed in \Cref{subsec:policy_action} increasing the number $n$ of equities in the basket; we have evidence that the $(n-1)$-sphere parameterization  \eqref{eq:euler_angles} suffers of labelling problems when $n>2$. In fact if we consider a spherical case ($n=3$), \cref{eq:euler_angles} introduces an asymmetry in the action space (\Cref{fig:n_sphere_parameterization} on the left) that translates into the choice by the agent of sub-optimal policies during the training phase. Because of that we privilege the action parameterization through normalization \eqref{eq:normalized_action}. 
 In \Cref{fig:n_sphere_parameterization} on the right we provide the learning curve of a $5\times8$ ANN\footnote{We use the notation $L\times N$ to describe a FFNN with $L$ hidden layers each made up of $N$ neurons.} with $\tanh$ as activation function, $\gamma=1$, $c_1=0.7$ and $c_2=0$. This learning curve, just like all the ones we will show in this work, is the best result in term of reward on the last $10^7$ episodes among four simulations with different initial seed for the weights $\theta$. This is done since the objective function is not convex.

@@

\begin{figure}[ht!]
	\centering
	\includegraphics[height=6cm,width=7.2cm]{Activation_functions_Learning_curve_1x40_network_confront_activation_strategy_EUR_8x5_900dpi.pdf}
	\includegraphics[height=6cm,width=7.2cm]{Value_function_Learning_curve_1x40_beta_confront_varying_strategy_EUR_8x5_400dpi_compressed.pdf}
	\includegraphics[height=6cm,width=7.2cm]{Hidden_Layers_Learning_curve_3asset_varystrategy_window1000000_EUR_5e8timestep_comparison_900dpi_compressed.pdf}
	\caption{Fine tuning simulations: on the top-left learning curves with different activation functions, on the top-right grid search on the policy-value trade-off and on bottom-centre learning curves on ANNs architectures. On the horizontal axis the number of training episodes. The solid lines are the moving average of the realized rewards on the last $5\times 10^5$ episodes. The shadows represent the $98\%$ confidence intervals. The dash-dotted horizontal red lines are the conservative TVO prices computed according to the optimal risky portfolio composition through \cref{eq:bs_closed_solution}.}
	\label{fig:fine_tuning}
\end{figure}  
\clearpage
\subsection{Local Volatility Dynamics}
In this section we study the TVS control problem assuming a local volatility model for the dynamics of the risky assets. Thus in this case we have a diffusive term in \Cref{eq:Equity_process} that is a deterministic function both of time and state, i.e. the spot price, $\nu_t=\nu(t,S_t)$. This additional dependency of volatility makes the problem of finding the optimal strategy non-trivial; in fact if we consider the whole equity smiles then the second order term in the HBJ equation \eqref{eq:HJB_general} is non null and thus a closed formula for $\alpha^*$ is not available anymore. Because of that one must resort on numerical techniques to recover the problem solution. Unlike BS model where the strategy depends only on time, in LV dynamics there are no unnecessary information provided by the state block in \Cref{eq:state} to take the optimal action.

@@

\section{Conclusion and Further Developments}
In this paper we described a non-trivial control problem related to derivative contracts on target volatility strategies. We considered a bank selling a call option to a fund manager as protection  on the capital
invested on the TVS. We showed how the presence of different funding costs coming from hedging the risky assets underlying the TVS, makes the bank solving a stochastic optimal control problem to adjust the price of the protection, since the bank strategy is not self-financing. We derived a closed form solution of the control problem in a Black and Scholes framework in two different ways: first by applying the Gy\"ongy  Lemma and then by writing the Hamilton-Jacobi-Bellman equation. We studied the problem both in the BS world and in the local volatility one where the solution is not available and thus numerical investigation is needed. We tackled the problem by means of the novel Reinforcement Learning techniques, such as the proximal policy optimization. We used the BS model, where the solution is \textit{a priori} known, as benchmark to perform a series of fine-tuning of the RL algorithm hyper-parameters, such as the artificial neural network architecture. Then we investigated the problem under LV dynamics. From our simulations we had evidence that under this model the RL agent learnt as optimal strategy the application path-wise of the BS solution. This topic is still under study since our results for the LV model do not provide a formal demonstration of that the problem solution is the path-wise BS one. For this reason the natural developments of this work is to compare our RL results with other standard techniques like the least-square Monte Carlo. If the resulting comparison is positive, then it will provide robustness of our LV results.
\begin{comment}
